---
title: "DSC 680 Project 1"
author: "Joshua Gardner"
date: "September 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Up

The first thing I need to do is start setting up this project. I have the data collected, but I need to now import the data into *R*.

```{r}
flu_by_week <- read.csv("C:/Users/yasam/OneDrive/Documents/Grad School/DSC680 Applied Data Science/Project 1/Data/Flu Age and Virus Breakdowns/VirusViewByWeek.csv", header=TRUE)
head(flu_by_week)
flu_national <- read.csv("C:/Users/yasam/OneDrive/Documents/Grad School/DSC680 Applied Data Science/Project 1/Data/Flu National Level/WHO_NREVSS_Combined_prior_to_2015_16.csv", header=TRUE)
head(flu_national)
measles <- read.csv("C:/Users/yasam/OneDrive/Documents/Grad School/DSC680 Applied Data Science/Project 1/Data/Measles/Number of Measles Cases Reported by Year.csv", header=TRUE)
head(measles)
whooping_22_15 <- read.csv("C:/Users/yasam/OneDrive/Documents/Grad School/DSC680 Applied Data Science/Project 1/Data/Whooping Cough/Pertussis Cases by Year 1922-2015.csv", header=TRUE)
head(whooping_22_15)
whooping_by_age <- read.csv("C:/Users/yasam/OneDrive/Documents/Grad School/DSC680 Applied Data Science/Project 1/Data/Whooping Cough/Pertussis Incidence by Age Group and Year 1990-2017.csv", header=TRUE)
head(whooping_by_age)
```

Alright, I have the initial data imported. Now I need to get the data put together into a format that I can use to compare the different diseases. The easiest to use would be to get the total number of cases for each year for each disease and combine them into a single database.

Sadly, I wasn't able to find enough data to include the bubonic plague in this analysis. The most I can say is that there have been confirmed cases of the bubonic plague within the US in the past few years.

## Cleaning the Data

To start cleaning the data, I will need to compile each set of data into the total number of cases for each year. I wish I could have found weekly data for each of the diseases, but I couldn't get that data for each disease. But getting back to starting to clean the data.

### Flu

```{r}
head(flu_national)
```

The interesting thing about the flu is that there are multiple strains of flu. I can look at the different strains of flu.

```{r, include=FALSE}
library(plyr)
library(dplyr)
```

```{r}
h3n2 <- flu_national %>% group_by(YEAR) %>% summarize(H3N2v = sum(H3N2v))
h1n1 <- flu_national %>% group_by(YEAR) %>% summarize(H1N1 = sum(A..2009.H1N1.))
h1 <- flu_national %>% group_by(YEAR) %>% summarize(H1N1 = sum(A..H1.))
h3 <- flu_national %>% group_by(YEAR) %>% summarize(H1N1 = sum(A..H3.))
subtype_not_pref <- flu_national %>% group_by(YEAR) %>% summarize(na_sub = sum(A..Subtyping.not.Performed.))
unable_to_subtype <- flu_national %>% group_by(YEAR) %>% summarize(unable = sum(A..Unable.to.Subtype.))
flu_cases_df <- data.frame(h3n2, h1n1[2], h1[2], h3[2], subtype_not_pref[2], unable_to_subtype[2])
cols <- c('year','h3n2', 'h1n1', 'h1', 'h3', 'subtype_not_pref', 'unable_to_subtype')
colnames(flu_cases_df) <- cols
head(flu_cases_df)
```

### Measles

Now that the number of flu cases have been grouped by year for each different strain of flu, it's time to move onto getting the Measles data into the same format. Fortunately, the data is already in this format, so no further work is needed right now.

### Whooping Cough

The data for the Whooping Cough is also already in a yearly format.

### Final Step

The final step is to combine all three diseases into a single data frame. What makes this challenging is that each disease has a different time frame that it is looking at. There are different approaches that can be taken to complete this. I will be combining the data to contain all of the available data, which means there will be a lot of `null` values that I will need to consider when doing the analysis. If needed, I may restrict the data frame to only contain the years for which I have full data available, but I do want to explore all of the available data for each disease independently.

```{r}
colnames(measles) <- c('year', 'measles')
colnames(whooping_22_15) <- c('year', 'whooping_cough')
mid_step <- join(whooping_22_15, measles, by='year', type='full')
diseases_full <- join(mid_step, flu_cases_df, by='year', type='full')
mid_step <- join(whooping_22_15, measles, by='year', type='inner')
diseases_inner <- join(mid_step, flu_cases_df, by='year', type='inner')
```

And here we have the combined data, both the `full` and the `inner` joins. However, there are a few issues with these dataframes. For example:

```{r}
str(diseases_full)
```

The `whooping_cough` variable is a factor, not integers. So let's correct that. What's causing the problem is that there are commas in the numbers... Hmmm, the easies fix would be to just remove the commas from the base file. Let's just do that, which would then make this whole section not make any sense. Oh well.

Now that we've corrected the factor issue, we can see that everything is imported as integers. Now I can start exploring this data.

## Exploring the Data

The next thing to do is to start looking at the data.

### Histograms

One of the first things I want to look at the different histograms.

```{r, include=FALSE}
library(ggplot2)
```

```{r, echo=FALSE, warning=FALSE}
ggplot(diseases_full, aes(x=whooping_cough)) + geom_histogram(bins=30)
ggplot(diseases_full, aes(x=measles)) + geom_histogram(bins=30)
ggplot(diseases_full, aes(x=h3n2)) + geom_histogram(bins=30)
ggplot(diseases_full, aes(x=h1n1)) + geom_histogram(bins=30)
ggplot(diseases_full, aes(x=h1)) + geom_histogram(bins=30)
ggplot(diseases_full, aes(x=h3)) + geom_histogram(bins=30)
```

Hmmm, none of the diseases looks to be very normally distributed. Looking at the different histograms, there do appear to be some outliers within the data. However, these outliers of of interest to my analysis, so I will be keeping these points of data in my analysis.

### ScatterPlots

The next thing I want to look at are the different scatterplots. Just as a side note, this is still just the exploration stage, not the visualization stage. These initial visualizations are going to be rough.

```{r}
plot(flu_cases_df)
plot(diseases_full)
plot(diseases_inner)
```

Hmm, these are interesting to look at. But how about the basic correlations?

```{r}
# Correlations of the different strains of Flu
cor(flu_cases_df)

# Correlations of the full join of the different diseases. But we need to deal with the NA values. Let's try to use pairwise.complete.obs
cor(diseases_full, use='pairwise.complete.obs')

# Correlations of the inner join of the different diseases
cor(diseases_inner)
```

Hmmm, there are a lot of interesting things here. But let's discuss this in a bit in the next section. Before jumping there, let's actually produce some plots of each disease over the years. Then look at some of how the different disease compare to each other.

#### Diseases by Year

```{r}
plot(diseases_full$year, diseases_full$whooping_cough, type='o')
plot(diseases_full$year, diseases_full$measles, type='o', xlim=c(1995,2020))
plot(diseases_full$year, diseases_full$h3n2, type='o', xlim=c(1995,2020))
plot(diseases_full$year, diseases_full$h1n1, type='o', xlim=c(1995,2020))
plot(diseases_full$year, diseases_full$h1, type='o', xlim=c(1995,2020))
plot(diseases_full$year, diseases_full$h3, type='o', xlim=c(1995,2020))
```

Here we can see that most of these diseases have had an outbreak within the past few years. The H1 strain of the flu has not had many cases over the past few years. Some are still trending up. Looking at these graphs, I'm tempted to say that the h3 strain of the flu or the measles is the most likely to have the next outbreak.

Now having said that, the two diseases would have different impact numberwise for an outbreak. What I mean by that is that an outbreak of measles has historically stayed under 1,000. On the other hand, the h3 strain of the flu has been in the 20,000's in recent years.

If we were only looking at the number of those infected with each disease, the number of measles would be drowned out by the other diseases. It's a good thing that these are tracked differently.

#### Measles

I am curious how the number of measles cases look when compared to the other diseases. Let's take a look at these.

```{r}
plot(diseases_full$measles, diseases_full$whooping_cough, ylim=c(0,60000), xlim=c(0,800))
plot(diseases_full$measles, diseases_full$h3n2, xlim=c(0,800))
plot(diseases_full$measles, diseases_full$h1n1, ylim=c(0,40000), xlim=c(0,800))
plot(diseases_full$measles, diseases_full$h1, ylim=c(0,10), xlim=c(0,800))
plot(diseases_full$measles, diseases_full$h3, xlim=c(0,800))
```

Interesting. The measles looks to be positively correlated to all of the other diseases except for the h3n2 and h1 strains of the flu. Looking at the calculated correlations above, the measles is also weakly correlated to the whooping cough.

#### Whoopping Cough

Let's now take a look at the whooping cough

```{r}
plot(diseases_full$whooping_cough, diseases_full$measles, xlim=c(0,50000), ylim=c(0,800))
plot(diseases_full$whooping_cough, diseases_full$h3n2, xlim=c(0,50000))
plot(diseases_full$whooping_cough, diseases_full$h1n1, xlim=c(0,50000))
plot(diseases_full$whooping_cough, diseases_full$h1, xlim=c(0,50000))
plot(diseases_full$whooping_cough, diseases_full$h3, xlim=c(0,50000))
```

Hmmmm, the whooping cough is positively correlated with the h3n2, h1n1, and h3 strains of the flu, and negatively correlated with the measles and h1 strain of flu. This honeslty surprises me. I thought that the h1 and h1n1 strains would have had similar correlations with the other diseases.

## Correlations

Now that I've explored the data, let's take a look at the correlations between the different diseases. Above, I ran Pearson's correlation test. However, not all of the variables are normal. That is to say that each of the diseases are skewed right.

This means that Pearson's Correlation test is not the ideal test to use. Pearson's correlation test assumes that the data is normally distributed. A better correlation test to use in this case would be Spearman's Correlation test. When using Spearman's Correlation test, the data does not need to be normally distributed, but the data must be at least ordinal. Here are the correlation values using Spearman's test:

```{r}
cor(diseases_inner, method='spearman')
```

There are a couple of interesting things in these correlations. For one, all of the diseases are positively correlated with the year, except for the h1 strain of the flu. This implies that all of these diseases are becoming more prevalent as time goes on, except for h1 flu.

Something else to note is that the h1 strain of flu is negatively correlated with every other disease that we've looked at in this analysis. I was not expecting to see that.

The strongest correlation is between the whooping cough and the h3n2 strain of flu at `0.8406680`.

One last thing is that the number of cases of flu that are not subtyped is also positively correlated with the year, implying that there are more and more cases of the flu that are not classified by strain.

## Covariate Relationships

Let's start looking at how each of the diseases relate to the year.

### Whooping Cough

```{r, warning=FALSE}
reg1 <- lm(whooping_cough ~ year, diseases_full)
summary(reg1)
ggplot(diseases_full, aes(x=year, y=whooping_cough)) + geom_point() + geom_smooth(method='lm')
```

Hmmm, this is the analysis from 1925 on, but what about more recently?

```{r, warning=FALSE}
d2000 <- filter(diseases_full, year >= 2000)
reg1 <- lm(whooping_cough ~ year, d2000)
summary(reg1)
ggplot(d2000, aes(x=year, y=whooping_cough)) + geom_point() + geom_smooth(method='lm')
```

Hmmm, interesting. Fromm 2000 - 2017, the number of cases of whooping cough have been increasing. Overall, since the vaccine for the whooping cough was developed, the number of cases each year has dropped dramatically, but it is on the rise again.

### Measles

```{r, warning=FALSE}
reg1 <- lm(measles ~ year, diseases_full)
summary(reg1)
ggplot(diseases_full, aes(x=year, y=measles)) + geom_point() + geom_smooth(method='lm')
```

This gives us a squashed view, so let's expand that out.

```{r, warning=FALSE}
reg1 <- lm(measles ~ year, d2000)
summary(reg1)
ggplot(d2000, aes(x=year, y=measles)) + geom_point() + geom_smooth(method='lm')
```

Hmmm, the interesting thing here is that the regression line is not statistically significant. The p-value 0.0979. However, it can still be seen that the number of cases is increasing.

### h3n2 Strain of Flu

```{r, warning=FALSE}
reg1 <- lm(h3n2 ~ year, diseases_full)
summary(reg1)
ggplot(diseases_full, aes(x=year, y=h3n2)) + geom_point() + geom_smooth(method='lm')
```

Once again, let's focus in on what data we do have for h3n2.

```{r, warning=FALSE}
reg1 <- lm(h3n2 ~ year, d2000)
summary(reg1)
ggplot(d2000, aes(x=year, y=h3n2)) + geom_point() + geom_smooth(method='lm')
```

This is another disease that is not statistically significat (p-value = 0.2924). Having said that, this disease had an outbreak in 2012 after having 12 years of no cases. There was a single case reported in 2015, so the disease is tappering off again. The projected increase is not significant without more data.

### h1n1 Strain of Flu

```{r, warning=FALSE}
reg1 <- lm(h1n1 ~ year, diseases_full)
summary(reg1)
ggplot(diseases_full, aes(x=year, y=h1n1)) + geom_point() + geom_smooth(method='lm')
```

Like with the earlier regressions, let's zoom in on this one.

```{r, waring=FALSE}
reg1 <- lm(h1n1 ~ year, d2000)
summary(reg1)
ggplot(d2000, aes(x=year, y=h1n1)) + geom_point() + geom_smooth(method='lm')
```

Interesting. This is another case of the regression line not being statistically significant (p-value = 0.4092). Part of that will be because of the outbreak in 2009.

### h1 Strain of Flu

```{r, warning=FALSE}
reg1 <- lm(h1 ~ year, diseases_full)
summary(reg1)
ggplot(diseases_full, aes(x=year, y=h1)) + geom_point() + geom_smooth(method='lm')
```

Let's also zoom in on this one.

```{r, warning=FALSE}
reg1 <- lm(h1 ~ year, d2000)
summary(reg1)
ggplot(d2000, aes(x=year, y=h1)) + geom_point() + geom_smooth(method='lm')
```

Here's another interesting case where the regression is not statistically significant (p-value = 0.6831). This is due to the outliers in the data. What I find really interesting here, though, is that the number of cases seems to be decreasing year over year for the h1 strain of flu.

### h3 Strain of Flu

```{r, warning=FALSE}
reg1 <- lm(h3 ~ year, diseases_full)
summary(reg1)
ggplot(diseases_full, aes(x=year, y=h3)) + geom_point() + geom_smooth(method='lm')
```

Let's also look closer at this.

```{r, warning=FALSE}
reg1 <- lm(h3 ~ year, d2000)
summary(reg1)
ggplot(d2000, aes(x=year, y=h3)) + geom_point() + geom_smooth(method='lm')
```

Oh, now this is interesting. Where most of the previous regression lines were not statistically significant, this one is (p-value = 7.877e-05). This means that the number of cases of h3 flu have been increasing and are likely to continue increasing year over year. That is worrisome.

## Predictions

Before jumping into the predictions, I want to summarize the different regressions quickly.

Disease        | Change   | p-value |Significant?
---------------|----------|---------|------------
Whooping Cough |Increasing|0.02667  |Yes
Measles        |Increasing|0.0979   |No
h3n2 Flu       |Increasing|0.2924   |No
h1n1 Flu       |Increasing|0.4092   |No
h1 Flu         |Decreasing|0.6831   |No
h3 Flu         |Increasing|7.887e-05|Yes

The most alarming thing is that most of these diseases are projected to increase in the number of cases year over year. Having said that, only the Whooping Cough and the h3 strain of the flu were statistically significant. This isn't to say that the other diseases will not increase, just that more data is needed before this can be concluded with certainty.

When looking at the different graphs and predictions, the three diseases that are predicted to have the most cases in the next few years are the Whooping Cough, h1n1 Flu, and h3 Flu. What's interesting is that both the Whooping Cough and h3 Flu predictions are statistically significant, but the h1n1 Flu prediction is not.

Based on the predictions, the h1n1 Flu will have the most infections in 2019, but the confidence interval allows for a wide varitability within the prediction. The predictions for the Whooping Cough have a tighter confidence interval and is statistically significant. Because of this, I predict that the Whooping Cough will have the most cases within the next few years. This is depressing as the vaccine has been highly effective against the whooping cough historically.

## Addressing Research Questions

1. Which diseases are strongly correlated to each other?
    + Whooping Cough and h3n2 Flu (0.87918124)
    + Measles and h1n1 Flu (0.84685604)
    + Measles and h3 Flu (0.54350205)
    + h1 Flu and h3 Flu (-0.83425512)
2. Do preventable diseases have outbreaks during the same time periods?
    + No, the different diseases have outbreaks during different years. Additionally, the outbreaks of each disease are of differing numbers. For example, an outbreak of the Measles could be a few hundred cases in a year. But an outbreak of the h1n1 Flu is 97,258 cases in a single year.
    + Some of the highly correlated diseases do fluctuate together, but very few have local maximums during the same year.
3. How has the current anti-vaxx movement impacted the number of infected cases?
    + Sadly, I was not able to determine the impact the anti-vaxx movement has on the number of infections. Finding the number of people who have been vacinated vs those who have not been is not readily available to the public, and is not available on a national level.
    + It can be tempting to say that the anti-vaxx movement has influenced the current trend of the number of infections increasing, but this analysis did not attempt to determine the impact of the anti-vaxx movement.
4. Looking at past trends in the numbers of infected individuals, what are the projected number of infected individuals over the next few years?
    + For five of the six diseases that were analyized, the number of infections are projected to increase year over year. However, only 2 of the projections are statistically significant.
    + Only the h1 Flu is projected to decrease over the next few years. Having said this, though, this projection is not statistically significant.
5. Looking at each disease's projected infections, which disease is likely to have the next outbreak?
    + Five diseases are projected to continue increasing in the number of infections as time goes on. Of the five, the Whooping Cough, h1n1 Flu, and h3 Flu are projected to have the highest number of infections.
    + What constitutes an outbreak is different for each disease. For example, an outbreak of the Measles would be a few hundred cases per year, but an outbreak of the Whooping Cough is in the tens of thousands per year.
    + Out of all of the diseases here, the disease that is most likely to have the highest number of infections over the next few years is the Whooping Cough.
    
## Summary

Over the past few years, more and more news stories are focusing on which disease is currently in the midst of an outbreak. Looking at the national infection numbers for six different diseases, it is possible to predict which diseases will have a high number of new infections over the next few years and which of the diseases are correlated to each other.

Each disease that was analyzed has a different threshold for what an outbreak looks like. Because of this, each disease has had an outbreak over the past few years and has the chance of having another outbreak within the next few years. 

Of the analyzed diseases, the Whooping Cough, h1n1 Flu, and h3 Flu are all projected to have high numbers of infections over the next few years. Of these diseases, the Whooping Cough is the most likely to have the highest number of infections.

## Extras

This portion is just to generate some more graphs for the final PowerPoint.

```{r}
ggplot(diseases_full, aes(x=year, y=whooping_cough)) + geom_smooth(method='loess') + geom_point() + theme_minimal()
ggplot(diseases_full, aes(x=year, y=measles)) + geom_smooth(method='loess') + geom_point() + theme_minimal() + xlim(c(2009, 2020))
ggplot(diseases_full, aes(x=year, y= h3n2)) + geom_smooth(method='loess') + geom_point() + theme_minimal() + xlim(c(2000, 2020))
ggplot(diseases_full, aes(x=year, y=h1n1)) + geom_smooth(method='loess') + geom_point() + theme_minimal() + xlim(c(2000, 2020))
ggplot(diseases_full, aes(x=year, y=h1)) + geom_smooth(method='loess') + geom_point() + theme_minimal() + xlim(c(2000, 2020))
ggplot(diseases_full, aes(x=year, y=h3)) + geom_smooth(method='loess') + geom_point() + theme_minimal() + xlim(c(2000, 2020))
```